---
title: "Serve weather events impact on public health and economy of United States"
author: "Hoang An"
date: "5/3/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

This project involves exploring the U.S. National Oceanic and
Atmospheric Administration's (NOAA) [storm
database](https://www.ncdc.noaa.gov/stormevents/) This database tracks
characteristics of major storms and weather events in the United States,
including when and where they occur, as well as estimates of any
fatalities, injuries, and property damage.

For preventing such outcomes to the extent possible is a key concern.
There are 2 questions which this report would like to address: 
  1. Across the United States, which types of events are most harmful with respect
to population health? 
  2. Across the United States, which types of events have the greatest economic consequences?

## Data processing

### Load data from zip file

```{r include=FALSE}
# Load popular package to work with big dataset
  library(dplyr)
  library(ggplot2)
  library(lubridate)        
```

The data for this analysis come in the form of a comma-separated-value
file compressed via the bzip2 algorithm to reduce its size. The source
data file is downloaded from [this
link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

```{r}
# Load data file into R
  StormData <- read.csv("C:/Users/Admin/Desktop/RStudio_et_Github/Reproducible Research/Final assigment/repdata_data_StormData.csv.bz2")

# Summarize the loaded data frame
  str(StormData)
```

### Subset data

-   Explanation of column names exits in [this
    link](https://cran.r-project.org/web/packages/noaastormevents/vignettes/details.html%5D).
    However, for the scoop of this analysis, only public heath and
    economy related variables are subset, including:

    -   `BGN_DATE` Beginning date
    -   `EVTYPE` Weather event types
    -   `FATALITIES` and `INJURIES` People die or injured by the event,
        which are impact on public health
    -   `PROPDMG` and `CROPDMG` Property damage and crop damage, which
        are impact on public health
    -   `PROPDMGEXP` and `CROPDMGEXP` Unit (by USD) for property and
        crop damage\

-   According to the NOAA
    (<https://www.ncdc.noaa.gov/stormevents/details.jsp>), only since
    1996, they can record all type of events. For comparison between
    events, data older than the year 1996 should be eliminated

-   Any observations with NA values should be eliminated

```{r}
# Subset needed data
  data <- StormData %>% select(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) 

# Reformat and select data since 1996
  data$BGN_DATE <- as.Date(data$BGN_DATE, "%m/%d/%Y")
  
# Filter data 
  data <- data %>% filter(BGN_DATE >= "1996/01/01")  # Data since 1996
  data <- data %>% filter (!is.na(data))    # Drop NA value
  data <- data %>% select(-BGN_DATE)        # Drop time data
```
### Process unit for Property and Crop damage

- Variables PROPDMGEXP
and CROPDMGEXP are unit (in USD) for PRODMG and CROPDMG, respectively.
However, their value input was coded as: 
    + "", "?", "+", "-": 1 
    + "0": 1
    + "1": 10 
    + "2": 100 
    + "3": 1.000 
    + "4": 10.000 
    + "5": 100.000 
    + "6": 1.000.000
    + "7": 10.000.000 
    + "8": 100.000.000 
    + "9": 1.000.000.000 
    + "H": 100 
    + "K": 1.000
    + "M": 1.000.000 
    + "B": 1.000.000.000 
    
  => For even unit, all the code in character should be transfer into number

```{r}
# For PROPDMGEXP
  ## Check for code using
    table(data$PROPDMGEXP)

  ## => Code appear in this column include 0, B, K, M

  ## Replace for suitable value for 0, B, K, M
  data$PROPDMGEXP <- gsub("0","1",data$PROPDMGEXP)
  data$PROPDMGEXP <- gsub("B","9",data$PROPDMGEXP) 
  data$PROPDMGEXP <- gsub("K","3",data$PROPDMGEXP) 
  data$PROPDMGEXP <- gsub("M","6",data$PROPDMGEXP)
  data$PROPDMGEXP <- as.numeric(data$PROPDMGEXP)
  
# For CROPDMGEXP
  ## Check for code using
    table(data$CROPDMGEXP)

  ## => Code appear in this column include B, K, M
    
  ## Replace for suitable value for B, K, M
  data$CROPDMGEXP <- gsub("B","9",data$CROPDMGEXP) 
  data$CROPDMGEXP <- gsub("K","3",data$CROPDMGEXP) 
  data$CROPDMGEXP <- gsub("M","6",data$CROPDMGEXP)
  data$CROPDMGEXP <- as.numeric(data$CROPDMGEXP)
  
# Null value will return as NA, they are the value without unit, therefore we can ignore them
    
```

-   Calculating economy impact

```{r}
  data[,"PROPERTY"] <- with(data,PROPDMG*10^PROPDMGEXP)
  data[,"CROP"] <- with(data,CROPDMG*10^CROPDMGEXP)
```

## Analysis

### Calculate health impact

Considering number of death and injury people have the same weight
impact on public health. Note that there is many event types, however,
we will only look at events with top highest impact.

```{r,echo=FALSE}
# PH is a new variable represent pubic health impact
  Health <- data %>% mutate(data,PH = FATALITIES + INJURIES) 
```

```{r}
# Calculate health impact by events 

  Health <- aggregate(Health$PH,by=list(Health$EVTYPE),sum,na.rm=TRUE)
  Health <- subset(Health,x>quantile(x,prob=0.98))
  Health <- Health[order(-Health$x),]
  colnames(Health) <- c("Event","Impact")
  Health 
```

### Calculate economy impact

Considering property damage and crop damage have the same weight impact
on economy

```{r}
#EC is a new variable represent economy impact
  Economy <- data %>% mutate(data,EC = PROPERTY + CROP) 
```

```{r}
# Calculate health impact by events 

  Economy <- aggregate(Economy$EC,by=list(Economy$EVTYPE),sum,na.rm=TRUE)
  Economy <- subset(Economy,x>quantile(x,prob=0.98))
  Economy <- Economy[order(-Economy$x),]
  colnames(Economy) <- c("Event","Impact")
  Economy
```
## Results
### Events with biggest impact on public health

```{r}
# Take 10 biggest impact for plotting
Health <- head(Health,10)

# Plot Event by impact

health.plot <- ggplot(Health, aes(x = Event, y = Impact, fill = Event)) +
              geom_bar(stat = "identity") +
              coord_flip() +
              ylab("Total number of health impact") +
              ggtitle("Weather event types impact on public health") +
              theme(plot.title = element_text(hjust = 0.5))

print(health.plot)
```


According to the graph, Tornado is the event with biggest impact on public health (both fatalities and injuries)

### Events with biggest impact on economy

```{r}
# Take 10 biggest impact for plotting
Economy <- head(Economy,10)

# Plot Event by impact

economy.plot <- ggplot(Economy, aes(x = Event, y = Impact, fill = Event)) +
              geom_bar(stat = "identity") +
              coord_flip() +
              ylab("Total number of economy impact by USD") +
              ggtitle("Weather event types impact on economy") +
              theme(plot.title = element_text(hjust = 0.5))

print(economy.plot)
```


According to the graph, Flood is the event with biggest impact on Economy (both property and crop)
